Seasonality in governing dynamics at submesoscales

### 2016-06-10 writeup/rocha_etal.tex
Bills small editorial fixes

Inertia-gravity waves, on the other hand, peak in late Summer/early Fall,
when the upper ocean is strongly stratified. Thus there exists a
strong seasonal modulation in the dominant upper-ocean submesoscale dynamics:
submesoscale turbulence dominates the upper-ocean dynamics in late Winter/early Spring,
whereas quasi-linear inertia-gravity waves prevail in late Summer/early Fall.
Because submesoscale turbulence is severely diminished in late Summer/early Fall,
the present results suggest that inertia-gravity waves account for most of the
summertime submesoscale SSH variability.

@@

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have suggested strong submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a strong seasonal cycle: wintertime
enhance small-scale turbulence de-stratify the upper ocean, yielding mixed layers
as deep as 300m. In late Spring/early Summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
The stratification and its seasonality is well-captured by both LLC simulations:
the skillful comparisons of model stratification to Argo climatology lends confidence
to the present analysis (supplemental material).

\begin{figure*}[ht]
\begin{center}
\hspace{-1.25cm}\includegraphics[width=.6\textwidth]{figs/fig1_1.pdf}\\
\vspace{-.125cm}
\includegraphics[width=.65\textwidth]{figs/fig1_2.pdf}
 \caption{(a) The study region with the subregion where the LLC outputs are
          analyzed. Colors represent the topography and black lines are absolute
          dynamic topography contours every 0.1 m. LLC 4320 (1/48$^\circ$) snapshots of surface vorticity (b and c) and a transect
          of potential density at 165$^\circ$E (d and e). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

\section{Statistics of surface lateral velocity tensor}
To study the seasonality in the surface velocity, we calculate the lateral velocity tensor
\begin{equation}
\nabla_h \bf{u}_h = \left[\begin{matrix} u_x & u_y\\ v_x&v_y \end{matrix}\right] \,.
\end{equation}
The components of the lateral velocity tensor are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$, lateral rate of strain
$\alpha \equiv [(u_x-v_y)^2 + (u_y + v_x)^2]^{1/2}$, and horizontal divergence $\delta \equiv u_x + v_y$.
These diagnostics highlight the submesoscale structures in the flow
and their analysis hint on the governing dynamics in the upper ocean
\citep[e.g.,][]{capet_etal2008a,shcherbina_etal2013}.

@@

\section{Seasonality in governing submesoscale dynamics}
The overall results of figure \ref{fig2}c strongly suggest that the
summertime submesoscale surface variability stems from different dynamics than
then wintertime variability. To better understand these differences, we calculate
joint probability distributions (jPDF) of
vorticity-divergence, vorticity-strain, and vorticity-laplacian of sea-surface height.

The April vorticity-strain jPDF have a shape characteristic of submesoscale turbulence
 (see figures \ref{fig3}f-g). The alignment of vorticity and strain $\alpha \sim \zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, and residual velocities, although
 the vorticity skewness reduces 1.4 to 1.13 from hourly to daily-averaged.
 The April vorticity-divergence jPDFs, however, have a egg-like shape that which again highlight
 the strong skewness in vorticity. Thus wintertime submesoscale surface velocity
 is strongly dominated by submesoscale turbulence --- this is true even for
 superinertial submesoscale currents. The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-laplacian of SSH
(figure \ref{fig3}o); it is only the daily-averaged velocity that is approximately
in geostrophic balance (figure \ref{fig3}l). While there is certainly much to be
learned using quasi-geostrophic models to study submesoscale turbulence, claims that
the latter is dominantly in geostrophic balance \cite{sasaki_etal2014,callies_etal2016}
is overstated; it is only the daily-averaged fields that are to a large extent
in  geostrophic balance (the jPDF of daily-averaged vorticity vs. laplacian of SSH
is a ellipsis with large eccentricity with main axis tilted by 45$^\circ$; figure \ref{fig3}n).

The October vorticity-strain jPDF show much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
 vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
 about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
 that persist in Summer) and one with a very steep slope. The vorticity-divergence
 jPDF is significantly different than the April one. In particular, the jPDF
 for daily-averaged currents is still egg-shaped but hourly and residual are
 ellipses about $\zeta=0$, likely associated with inertia-gravity waves.
That the submesoscale dynamics in October is mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-laplacian of SSH for hourly and
residual fields
(figure \ref{fig3}p and \ref{fig3}r), which is an ellipsis aligned in the vertical axis.
Of course daily-averaging the model fields filter out the ageostrophic flows and
the daily-averaged flow is essentially geostrophic.

@@

\section{Summary and Conclusion}
Our main finding is that
near-surface upper-ocean inertia-gravity waves undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
the submesoscale turbulence. \cite{dasaro1978} showed that linear internal waves
velocity in the mixed-layer strongly depends on the density jump at the mixed-layer
base, with largest mixed-layer velocities when the jump is strongest.
In Summer, the shallow mixed-layer overlays a strong seasonal pycnocline,
and thus the internal waves projection onto the mixed-layer may be stronger
according to \cite{dasaro1978}'s arguments. An
alternative explanation is that the shape baroclinic modes experience significant
seasonal modulation, with significant summertime peak in their surface amplitude
 (supplemental material).
If the internal wave source has weak seasonal dependence, then a strong
seasonal cycle in the near-surface expression of internal waves is expected.
Internal tide generation does not show a seasonal modulation \citep[e.g.,][]{alford2003},
and therefore it is plausible to expect a strong seasonality in the the near-surface
expression of internal tides and
other small-scale internal waves generated through internal tide interactions.
Near-inertial wave generation peaks in Winter, but those waves project on large
horizontal scales \citep[e.g, ][]{qi_etal1995}.

Our work adds to recent studies that presented modeling \cite{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. We suggest that the dominance at inertia-gravity waves in
Summer is a consequence of both suppresion of submesoscale turbulence and
enhancement of inertia-gravity waves due to restratification of
the upper ocean. Of course, our study focuses on a single patch of ocean in the
vicinities of the Kuroshio Extension, which may be typical of mesoscale-rich
subtropical regions, but not
representative of other regions such as eastern boundary currents
and the middle of the subtropical gyre: care must be taken in overgeneralizing
these results.  We plan to report on the geographic variability of submesoscale
 seasonality  in a future study.

>>>>

Inertia-gravity waves, on the other hand, peak in late Summer/early Fall,
when the upper ocean is strongly stratified. Thus there exists a
strong seasonal modulation in the dominant upper-ocean submesoscale dynamics:
submesoscale turbulence dominates the upper-ocean dynamics in late Winter/early Spring,
whereas quasi-linear inertia-gravity waves prevail in late Summer/early Fall.
Because submesoscale turbulence is weakest in late Summer/early Fall,
the present results suggest that inertia-gravity waves account for most of the
summertime submesoscale SSH variability.

@@

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have suggested strong submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a strong seasonal cycle: wintertime
enhance small-scale turbulence de-stratify the upper ocean, yielding mixed layers
as deep as 300m. In late Spring/early Summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
The stratification and its seasonality is well-captured by both LLC simulations:
the skillful comparisons of model stratification to Argo climatology is necessary
for the analysis in this paper (supplemental material).
\begin{figure*}[ht]
\begin{center}
\hspace{-1.25cm}\includegraphics[width=.6\textwidth]{figs/fig1_1.pdf}\\
\vspace{-.125cm}
\includegraphics[width=.65\textwidth]{figs/fig1_2.pdf}
 \caption{(a) The study region with the subregion where the LLC outputs are
          analyzed. Colors represent the topography and black lines are absolute
          dynamic topography contours every 0.1 m. LLC 4320 (1/48$^\circ$) snapshots of surface vorticity (b and c) and a transect
          of potential density at 165$^\circ$E (d and e). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

\section{Statistics of surface lateral velocity tensor}
To study the seasonality in the surface velocity, we calculate the lateral velocity tensor
\begin{equation}
\nabla_h \bf{u}_h = \left[\begin{matrix} u_x & u_y\\ v_x&v_y \end{matrix}\right] \,.
\end{equation}
The components of the lateral velocity tensor are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$, lateral rate of strain
$\alpha \equiv [(u_x-v_y)^2 + (u_y + v_x)^2]^{1/2}$, and horizontal divergence $\delta \equiv u_x + v_y$.
These diagnostics highlight the submesoscale structures in the flow
\citep[e.g.,][]{capet_etal2008a,shcherbina_etal2013}.

@@

\section{Seasonality in governing submesoscale dynamics}
The overall results of figure \ref{fig2}c shows that summertime submesoscale surface
 variability stems from different dynamics than wintertime variability.
 To better understand these differences, we calculate
joint probability distributions (jPDF) of
vorticity-divergence, vorticity-strain, and vorticity-laplacian of sea-surface height.

The April vorticity-strain jPDF have a shape characteristic of submesoscale turbulence
 (see figures \ref{fig3}f-g). The alignment of vorticity and strain $\alpha \sim \zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, and residual velocities, although
 the vorticity skewness reduces 1.4 to 1.13 from hourly to daily-averaged.
 The April vorticity-divergence jPDFs, however, have a egg-like shape that which again highlight
 the strong skewness in vorticity. Thus wintertime submesoscale surface velocity
 is strongly dominated by submesoscale turbulence --- this is true even for
 superinertial submesoscale currents. The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-laplacian of SSH
(figure \ref{fig3}o); it is only the daily-averaged velocity that is approximately
in geostrophic balance (figure \ref{fig3}l). While there is certainly much to be
learned using quasi-geostrophic models to study submesoscale turbulence, claims that
the latter is dominantly in geostrophic balance
are overstated \citep{sasaki_etal2014,callies_etal2016}; it is only the daily-averaged fields that are to a large extent
in  geostrophic balance (the jPDF of daily-averaged vorticity vs. laplacian of SSH
is a ellipse with large eccentricity with main axis tilted by 45$^\circ$; figure \ref{fig3}n).

The October vorticity-strain jPDF show much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
 vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
 about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
 that persist in Summer) and one with a very steep slope. The vorticity-divergence
 jPDF is significantly different than the April one. In particular, the jPDF
 for daily-averaged currents is still egg-shaped but hourly and residual are
 ellipses about $\zeta=0$, likely associated with inertia-gravity waves.
That the submesoscale dynamics in October is mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-laplacian of SSH for hourly and
residual fields
(figure \ref{fig3}p and \ref{fig3}r), which is an ellipse aligned in the vertical axis.
Of course daily-averaging the model fields filter out the ageostrophic flows and
the daily-averaged flow is essentially geostrophic.

@@

\section{Summary and Conclusion}
Our main finding is that
near-surface upper-ocean inertia-gravity waves undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
the submesoscale turbulence. \cite{dasaro1978} showed that linear internal waves
velocity in the mixed-layer strongly depends on the density jump at the mixed-layer
base, with largest mixed-layer velocities when the jump is strongest.
In Summer, the shallow mixed-layer overlays a strong seasonal pycnocline,
and thus the internal waves projection onto the mixed-layer may be stronger
according to \cite{dasaro1978}'s arguments. An
alternative explanation is that the shape of the baroclinic modes significantly
change seasonally. In particular the surface amplitude of the baroclinic modes are
largest when the upper ocean is strongly stratified in late Summer/ early Fall
 (supplemental material).
If the internal wave source has weak seasonal dependence, then a strong
seasonal cycle in the near-surface expression of internal waves is expected.
Internal tide generation does not show a seasonal modulation \citep[e.g.,][]{alford2003},
and therefore it is plausible to expect a strong seasonality in the near-surface
expression of internal tides and
other small-scale internal waves generated through internal tide interactions.
Near-inertial wave generation peaks in Winter, but those waves project on large
horizontal scales \citep[e.g, ][]{qi_etal1995}.

Our work adds to recent studies that presented modeling \cite{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. We suggest that the dominance at inertia-gravity waves in
Summer is a consequence of both  of submesoscale turbulence and
enhancement of inertia-gravity waves due to restratification of
the upper ocean. Of course, our study focuses on a single patch of ocean in the
vicinities of the Kuroshio Extension, which may be typical of mesoscale-rich
subtropical regions, but not
representative of other regions such as eastern boundary currents
and the middle of the subtropical gyre: care must be taken in overgeneralizing
these results.  We plan to report on the geographic variability of submesoscale
 seasonality  in a future study.
